---
title: "NYC Food Scrap Drop Off"
author: "Justin Williams"
date: "3/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r results=FALSE, message=FALSE}
library(tidyverse)
library(dotenv)
library(RSocrata)
library(sf)
```

## Overview

In April 2020 <a href="https://www1.nyc.gov/assets/dsny/site/resources/press-releases/nyc-sanitation-announces-service-changes-related-to-covid-19-budget-cuts"> NYC Sanitation announced service changes</a> to their Curbside Composting program related to COVID-19 budget cuts. This meant any residents that were lucky enough to be able to partake in this greenhouse gas saving program, were cut off. Luckily some organizations rose to the challenge and began accepting food scrap drop-offs. For example in Greenpoint where I reside, <a href=" https://northbrooklynmutualaid.org/"> North Brooklyn Mutual Aid </a> started accepting food scraps in both <a href="https://www.nycgovparks.org/parks/msgr-mcgolrick-park"> Mcgolrick </a>, <a href="https://www.nycgovparks.org/parks/cooper-park"> Cooper </a> and <a href="https://www.nycgovparks.org/parks/transmitter-park"> Transmitter </a> parks. This enabled residents to minimize their environmental compost even when the city wasn't able to facilitate. As of August 2021 the Curbside Composting is back in effect, however it only services the following Community Boards:

<u>**Brooklyn**</u>
- 1, 2, 6 and 7

<u>**Manhattan**</u>
- 6 and 7

<u>**Bronx**</u>
- 8

**THIS IS EXTRA STUFF, NOT NEEDED FOR BASIC ANALYSIS**
```{r create-simple-df}
# create df to show current curbside composting program coverage
df_bb <- data.frame(boro_name = 
      c(rep("Brooklyn",4),"Bronx", rep("Manhattan",2), "Queens", "Staten Island"),
      cd = c(01,02,06,07,08,06,07,NA,NA))

# create empty col for boro number
df_bb$boro_num <- as.numeric(NaN)

# populate with boro numbers
df_bb[df_bb$boro_name == "Manhattan",]$boro_num <- 1
df_bb[df_bb$boro_name == "Bronx",]$boro_num <- 2
df_bb[df_bb$boro_name == "Brooklyn",]$boro_num <- 3
df_bb[df_bb$boro_name == "Queens",]$boro_num <- 4
df_bb[df_bb$boro_name == "Staten Island",]$boro_num <- 5


df_bb
```

City Council District shapefile was obtained from <a href="https://data.cityofnewyork.us/City-Government/City-Council-Districts/yusd-j4xi"> Open Data NYC </a>

````{r read-comm-b}
# read in community district shapefile
df_cd <- read_sf("./data/geo_export_f9ac259f-1dc1-4021-95bc-6fff4cc3fd89.shp")

head(df_cd)
```

I would like to join it with the made up data frame to show coverage of NYC Curbside Composting program.

```{r split-boro-cd}
df_cd <- df_cd %>% 
  mutate(boro = as.numeric(substr(df_cd$boro_cd,1,1)), 
         cd = as.numeric(substr(df_cd$boro_cd,2,3))) %>% 
  select(boro, cd, shape_area, shape_leng, geometry)

# preview df
head(df_cd)
```

Join data frame with Shapefile

```{r join-cd-df}
df_joined <- left_join(df_cd, df_bb, by = c("boro" = "boro_num", "cd" = "cd"))

head(df_joined)
```
Add names for missing borough names
```{r add-missing-boro-names}
df_joined[df_joined$boro == 1,]$boro_name <- "Manhattan"
df_joined[df_joined$boro == 2,]$boro_name <- "Bronx"
df_joined[df_joined$boro == 3,]$boro_name <- "Brooklyn"
df_joined[df_joined$boro == 4,]$boro_name <- "Queens"
df_joined[df_joined$boro == 5,]$boro_name <- "Staten Island"

head(df_joined)
```
```{r map-curbside}
plot(st_geometry(df_joined))
```


That said, not only are their a bunch of volunteer organizations that accept food scraps, but there are also a plethora of community gardens even some urban farms that will take your food scraps.

I wanted to do some exploratory data analysis to answer the following questions:

-   Which borough has the most drop-offs
-   Which Council Districts have the most drop-offs
-   What do those have in common?
-   What day/time is most popular for drop-off?
-   How many are open year round?
-   What do those have in common?
-   Do any Community Districts have more then one drop-off? What about Census Tracts?

The end result will be to develop a comprehensive understanding of where food-scrap drop off locations in NYC are located, and how the curbside composting program could work in collaboration with volunteer organizations to more equability offer its services.

For more information on why composting is good for the environment, check out this <a href= "https://youtu.be/wBLC-LOa7W8"> video </a>. 

## Data Dictionary

There are 17 columns in the data set.
See name, description and type below:

| Column Name 	| Description 	| Type 	|  	|
|---	|---	|---	|---	|
| Borough 	| NYC Borough where vendor is located. New York City’s boroughs are five county-level administrative divisions, with each one also being a state county. 	| Plain Text 	|  	|
| CouncilDist 	| NYC Council District Number. There are 51 Council districts throughout the five boroughs and is one is represented by an elected Council Member. 	| Number 	|  	|
| Food_Scrap_Drop_Off_Site_Name 	| Name of food scrap drop-off location 	| Plain Text 	|  	|
| Latitude 	| Latitude of food scrap drop-off location for mapping purposes. 	| Number 	|  	|
| Location 	| Street address or cross streets associated with food scrap drop-off location 	| Plain Text 	|  	|
| Longitude 	| Longitude of food scrap drop-off location for mapping purposes. 	| Number 	|  	|
| NTAName 	| Neighborhood Tabulation Area Name. Neighborhood Tabulation Areas are small area boundaries that were initially created by the Department of City Planning for small area population projections. However, NTAs are now being used to present data from the Decennial Census and American Community Survey. 	| Plain Text 	|  	|
| Notes 	| Additional site notes 	| Plain Text 	|  	|
| Object ID 	| An ObjectID is a unique, not null integer field used to uniquely identify rows in tables in a geodatabase. 	| Plain Text 	|  	|
| Operation_Day_Hours 	| Days of the week when food scraps can be dropped off. 	| Plain Text 	|  	|
| Open_Months 	| Months when food scraps can be dropped off at the location. 	| Plain Text 	|  	|
| Hosted_By 	| Name of the organization that services the food scraps that are dropped off. 	| Plain Text 	|  	|
| Website 	| Website associated with food scrap drop-off location. 	| Plain Text 	|  	|
| boroCD 	| Borough and Community District which is represented by a single-digit borough number followed by two-digit borough community district number. 	| Number 	|  	|
| ct2010 	| Census Tract (CT2010). The 2010 census tract in which the tax lot is located. 	| Plain Text 	|  	|
| point 	| Longitude and Latitude formatted for map "pin" 	| Point 	|  	|
| Precinct 	| Police Precinct associated with food scrap drop-off location 	| Number 	|  	|

## Read in data

<a href="https://opendata.cityofnewyork.us/> Open Data NYC </a> has a plethora of free and easily downloadable datasets. For our purposes I will be downloading the <a href ="https://data.cityofnewyork.us/Environment/Food-Scrap-Drop-Off-Locations-in-NYC/if26-z6xq"> Food Scrap Drop-Off Locations in NYC </a>. As of March 2022 this was last updated on November, 16th 2021, so I will be using the <a href="https://dev.socrata.com/connectors/rsocrata.html"> Socrata </a> API to get the data so it can easily be downloaded if updated, to update this analysis in the future.

First let's read in the data using the Socrata API:

```{r read-in-data-socrata-api, warning=FALSE}
# load api token
load_dot_env()

# read in data from api
df <- read.socrata("https://data.cityofnewyork.us/resource/if26-z6xq.json",
             app_token = Sys.getenv("SOCRATA_API"))

# convert to sf df
df_sf <- df %>% 
  st_as_sf(coords = c("latitude", "longitude"))

# delete uneeded geometry columns from original df
drops <- c("point.type", "point.coordinates")
df_sf <- df_sf[, !(names(df_sf) %in% drops)]

# save df to data folder so not to over use API
st_write(df_sf, paste('./data/Food_Scrap_Drop-Off_Locations_in_NYC',Sys.Date(),'.shp', sep = ""))
```

Re-load in data frame.
```{r load-data}
# if we don't load it in as a gdf doesn't load point coordinates.
df_sf <- read_sf("./data/Food_Scrap_Drop-Off_Locations_in_NYC2022-03-11.shp")

head(df_sf)
```
```{r}
# convert columns to numeric
cols.num <- c("councildist", "borocd", "precinct")
df_sf[cols.num] <- sapply(df_sf[cols.num], as.numeric)
```

## Analysis questions

So lets first focus on borough.

This will enable better visualizations with borough name and corresponding cd.

### Which Borough has the most drop-offs?

```{r boro-most-drop-offs}
df_sf %>% 
  count(borough) %>% 
  arrange(desc(n)) %>% 
  ggplot(aes(n, reorder(borough,n))) + 
    geom_bar(stat = "identity")
```
Looks like there are errors in the data, `Brookyln` is an incorrect spelling of `Brooklyn` and `Greenwich St. & Duane St.` is in Manhattan.

Correct these errors and redo code
```{r correct-borough-errors}
df_sf <- df_sf %>% 
  within(borough[borough == "Brookyln"] <- "Brooklyn") %>% 
  within(borough[borough == "Greenwich St. & Duane St"] <- "Manhattan")

# check unique boroughs
unique(df_sf$borough)
```
```{r redo-with-correct-data}
df_sf %>% 
  count(borough) %>% 
  arrange(desc(n)) %>% 
  ggplot(aes(n, reorder(borough,n), fill = borough)) + 
    geom_bar(stat = "identity") +
    scale_fill_discrete(guide = "none")
```

Now we get the true picture of counts per borough of food scrap drop-offs and can see Manhattan has the most followed by Brooklyn, Queens, Bronx and Staten Island.

**WOULD BE INTERESTING TO LOOK AT RATE to pop perhaps AS WELL IF HAVE TIME**

### Which Council Districts have the most drop-offs?

```{r cd-most-drop-offs}
df_sf %>% 
  group_by(borocd, borough) %>%
  summarise(
    count = n()
  ) %>% 
  arrange(desc(count)) %>%
  filter(count > 1) %>% 
  ggplot(aes(count, reorder(borocd,count), fill = borough)) +
    geom_bar(stat = "identity")
```

### Do any Council Districts have more then one drop off?

```{r cd-more-then-one}
df_sf %>%
  group_by(borocd, borough) %>% 
  summarise(
    count = n()
  ) %>% 
  filter(count >=5) %>% 
  arrange(desc(count))
```

#### What about Census Tracts?
```{r census-tract}
df_sf %>% 
  group_by(ct2010, borough) %>% 
  summarise(
    count = n()
  ) %>% 
  filter(count >= 2) %>% 
  arrange(desc(count))
```
What borough/neighborhood are the CT with more then one drop-off in?

```{r neighborhood-more-then-one}
df_sf %>% 
  group_by(ct2010, borough, ntaname) %>% 
  summarise(
    count = n()
  ) %>% 
  filter(count >= 2) %>% 
  arrange(desc(count))
```

#### What do these have in common?

### What day/time is the most popular for drop-ff?

```{r day-time}
head(df_sf$operation_day_hours)
```

### How many are open year round?
```{r open}
head(df_sf$open_months)
```

#### What do these have in common?
