---
+--title: "United States COVID-19 Cases and Deaths by State"
author: "Justin Williams"
date: "3/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE}
library(tidyverse)
library(dotenv)
library(RSocrata)
library(scales)
library(usmap)
library(lubridate)
```

## Overivew

Center for Disease Control and Prevention (CDC) reports aggregate counts of COVID-19 death and case number daily. Data is based on most recent numbers reported by states, territories and other jurisdictions. This depends on timely and accurate reporting, therefore there are some missing values. 

There are currently 60 public health jurisdictions reporting COVID-19 cases, including the 50 states, District of Columbia, NYC, American Samoa, Guam the Commonwealth of the Northern Mariana Islands, Puerto Rico and the US Virgin Islands. Additionally three independent countries in compacts of free association with the United States, Federated States of Micronesia, Republic of the Marshall Islands, and Republic of Palu.

```{r api-download-latest-data}
# load up hidden api key
load_dot_env()

# import dataset to R
df <- read.socrata("https://data.cdc.gov/resource/9mfq-cb36.json",
             app_token = Sys.getenv("SOCRATA_API"))

# save df to data folder
write_csv(df, paste("./data/us_covid_cases_and_deaths_",Sys.Date(),".csv", sep = ""))
```

```{r load-data, message=FALSE}
# load data saved
df <- read_csv("./data/us_covid_cases_and_deaths_2022-03-10.csv")
```

```{r data-preview}
head(df)
```

## Data dictionary 

There are 15 columns in this data set, you can see the titles, description and type in the data dictionary below:

| Column Name 	| Description 	| Type 	|
|---	|---	|---	|
| submission_date 	| Date of counts 	| Date & Time 	|
| state 	| Jurisdiction 	| Plain Text 	|
| tot_cases 	| Total number of cases 	| Number 	|
| conf_cases 	| Total confirmed cases 	| Number 	|
| prob_cases 	| Total probable cases 	| Number 	|
| new_case 	| Number of new cases 	| Number 	|
| pnew_case 	| Number of new probable cases 	| Number 	|
| tot_death 	| Total number of deaths 	| Number 	|
| conf_death 	| Total number of confirmed deaths 	| Number 	|
| prob_death 	| Total number of probable deaths 	| Number 	|
| new_death 	| Number of new deaths 	| Number 	|
| pnew_death 	| Number of new probable deaths 	| Number 	|
| created_at 	| Date and time record was created 	| Date & Time 	|
| consent_cases 	| If Agree, then confirmed and probable cases are included. <br> Not Agree, then only total cases are included. 	| Plain Text 	|
| consent_deaths 	| If Agree, then confirmed and probable deaths are included. <br>If Not Agree, then only total deaths are included. 	| Plain Text 	|

## Data Questions

Questions of interest are:

- Find the state with the highest and lowest deaths
- Compare the death rate both before and after the vaccine was released

## Question 1
### State with highest and lowest deaths

Let's first look at this from a pure count perspective

```{r get-death-count-state}
df_tot_deaths <- df %>% 
  select(submission_date, state, tot_cases, tot_death) %>% 
  group_by(state) %>% 
  filter(submission_date == max(submission_date)) %>% 
  ungroup()

# preview
head(df_tot_deaths)
```

Since `NYC` and `NYS` are reported separately, let's isolate these values.
May need to add them together to get a total and create a new row. 

```{r nys-andnyc}
df_tot_deaths[df_tot_deaths$state == 'NY' | df_tot_deaths$state == 'NYC',]
```

```{r graph-states with most deaths}
df_tot_deaths %>%
  arrange(desc(tot_death)) %>% 
  ggplot(aes(x = reorder(state, tot_death), y = tot_death)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  theme(legend.position = "none")
```


```{r top-10}
df_tot_deaths %>% 
  slice_max(tot_death, n = 10) %>% 
  ggplot(aes(reorder(state, tot_death), tot_death, fill = tot_death)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue")) +
    coord_flip()
    theme(legend.position = 'none')
```


```{r bot-10}
df_tot_deaths %>% 
  slice_min(tot_death, n = 10) %>% 
    ggplot(aes(reorder(state, -tot_death), tot_death, fill = tot_death)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue")) +
    coord_flip()
    theme(legend.position = 'none')
```

## Death rate by state

Now lets create a `death_rate` column and look at top_n that way.

```{r death-rate-by-state}
df_tot_deaths <- df_tot_deaths %>% 
  mutate(death_rate = tot_death / tot_cases)
```

Let's view top_n by rate

```{r top-10-rate}
df_tot_deaths %>% 
  slice_max(death_rate, n = 10) %>% 
  ggplot(aes(reorder(state, death_rate), death_rate, fill = death_rate)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue"), mid = "red", low = "white") +
    coord_flip()
    theme(legend.position = 'none')
```
```{r bot-10-rate}
df_tot_deaths %>% 
  slice_min(death_rate, n = 10) %>% 
    ggplot(aes(reorder(state, -death_rate), death_rate, fill = death_rate)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue"), mid = "red") +
    coord_flip()
    theme(legend.position = 'none')
```

# This section may have to be abandoned due to time constraints
## Simple map

Built is pacakge `usmaps` will suffice for this.
It will only have the 52 states but is fine for this purpose.

Have to sum `NY` and `NYC`
```{r sum-ny-nyc}
# sum total cases
NYS_tot_cases <- sum(df_tot_deaths$tot_cases[df_tot_deaths$state == "NY" | df_tot_deaths$state == "NYC"])

# sum total deaths
NYS_tot_deaths <- sum(df_tot_deaths$tot_death[df_tot_deaths$state == "NY" | df_tot_deaths$state == "NYC"])

# sum death rate
NYS_death_rate <- sum(df_tot_deaths$death_rate[df_tot_deaths$state == "NY" | df_tot_deaths$state == "NYC"])/2

#Latest date
dateLatest = as.POSIXct(max(df_tot_deaths$submission_date), origin = "1970-01-01", tz = "UTC")
dateLatest

# add new york summed row
rbind(df_tot_deaths, dateLatest, "NYS", NYS_tot_cases, NYS_tot_deaths, NYS_death_rate)

# remove separate ny and nyc
df_tot_deaths_ny <- subset(df_tot_deaths, state != "NYC" & state !="NY")

# rename NYS to NY
df_tot_deaths_ny <-  within(df_tot_deaths_ny, state[state == "NYS"] <- "NY")

df_tot_deaths
```

Join `df_tot_deaths_ny` with statepop

```{r join-with-statepop}
left_join(statepop, df_tot_deaths_ny, by = c("abbr" = "state"))
```


```{r simple-map-with-count}
# use usmap with built in pop
plot_usmap(data = statepop, values = "pop_2015", color = "blue") +
  scale_fill_continuous(low="white", high="red", name = "Population",
                        label = scales::comma)

statepop
```

## Question 2
### Compare death rate before and after vaccine being released

According to <a href="https://www.fda.gov/news-events/press-announcements/fda-takes-key-action-fight-against-covid-19-issuing-emergency-use-authorization-first-covid-19">FDA.gov</a>, the First COVID-19 Vaccine was authorized on **Dec 11th, 2020**. This was for emergency use only, however for simplicity sake we will use this as a place marker for this study. 

In this analysis we will use what the World Health Organization defines as <a href="https://www.who.int/news-room/commentaries/detail/estimating-mortality-from-covid-19">Infection fatality ratio (IFR)</a> as our death rate. T

This is calculated by:

$$\text{Infection fatality ratio (IFR, in %) = }\frac{\text {Number of deaths from  disease}}{\text{Number of infected  individuals}} \text{x 100}$$
They note that the true severity of a disease can be described by the Infection Fatality Ratio.

```{r before-vax}
df_before_vax <- df %>% 
  select(submission_date, state, tot_death, tot_cases) %>% 
  filter(submission_date <= '2020-12-11') %>%
  group_by(state) %>% 
  filter(submission_date == max(submission_date)) %>% 
  mutate(death_rate_before = tot_death / tot_cases) %>% 
  arrange(desc(death_rate_before))

# check for NaN
df_before_vax[is.na(df_before_vax$death_rate_before),]

# preview df
head(df_before_vax)
```
Missing values are simply divide by zero error, no need to account for them using na.rm = T will work.

```{r after-vax}
df_after_vax <- df %>% 
  select(submission_date, state, tot_death, tot_cases) %>% 
  filter(submission_date > '2020-12-11') %>%
  group_by(state) %>% 
  filter(submission_date == max(submission_date)) %>% 
  mutate(death_rate_after = tot_death / tot_cases) %>% 
  arrange(desc(death_rate_after))

# check for NaN
df_after_vax[is.na(df_after_vax$death_rate_after),]

# preview df
head(df_after_vax)
```
No missing values in `df_after_vax`

Overall before after vax.

```{r join-on-state}
df_death_rate <- df_before_vax %>% 
  left_join(df_after_vax, by = "state") %>% 
  select(state, death_rate_before, death_rate_after)

head(df_death_rate)
```

What is the total death rate before and after vaccine?

```{r total-death-rate-before-after}
# death rate before
before_vax <- sum(df_death_rate$death_rate_before, na.rm = T)/nrow(df_death_rate)

# death rate after
after_vax <- sum(df_death_rate$death_rate_after, na.rm = T)/nrow(df_death_rate)

# create new df
df_death_rate_sum <- data.frame(vax_status = c("before_vax", "after_vax"), 
                                death_rate = c(before_vax, after_vax))

df_death_rate_sum
```
```{r simple-death-rate-sum}
ggplot(df_death_rate_sum, aes(x=death_rate, y= vax_status, fill = vax_status)) +
  geom_bar(stat = 'identity')
```
## Conclusion

### Question 1

**Which states had the most and least deaths?**

- While California had the most deaths, the death rate (total cases / total deaths) was highest in NYC
- Least deaths in the continental United States was Vermont with ~600, they also were in the *bottom 10* for lowest death rate. Utah was first in lowest death rate with ~0.0045.

### Question 2

**What was the death rate before and after the vaccine was released?**

- If we use December 11th 2020 as the after vaccination date, the aggregate death rate for the entire data set is 0.016 before the vaccine was released, and 0.010 after. So we see a pretty significant decrease in total death rate after the vaccine was released. 







