---
title: "United States COVID-19 Cases and Deaths by State"
author: "Justin Williams"
date: "3/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE}
library(tidyverse)
library(dotenv)
library(RSocrata)
library(scales)
library(usmap)
library(lubridate)
```

## Overivew

Center for Disease Control and Prevention (CDC) reports aggregate counts of COVID-19 death and case number daily. Data is based on most recent numbers reported by states, territories and other jurisdictions. This depends on timely and accurate reporting, therefore some values are missing. 

There are currently 60 public health jurisdictions reporting COVID-19 cases, including the 50 states, District of Columbia, NYC, American Samoa, Guam, the Commonwealth of the Northern Mariana Islands, Puerto Rico and the US Virgin Islands. Additionally three independent countries in compacts of free association with the United States, Federated States of Micronesia, Republic of the Marshall Islands, and Republic of Palu.

```{r api-download-latest-data, warning=FALSE}
# load up hidden api key
load_dot_env()

# import dataset to R
df <- read.socrata("https://data.cdc.gov/resource/9mfq-cb36.json",
             app_token = Sys.getenv("SOCRATA_API"))

# save df to data folder
write_csv(df, paste("./data/us_covid_cases_and_deaths_",Sys.Date(),".csv", sep = ""))
```

```{r load-data, message=FALSE}
# load data saved
df <- read_csv(paste("./data/us_covid_cases_and_deaths_",Sys.Date(),".csv", sep=""))
```

```{r data-preview}
head(df)
```

## Data dictionary 

There are 15 columns in this data set, you can see the titles, description and type in the data dictionary below:

| Column Name 	| Description 	| Type 	|
|---	|---	|---	|
| submission_date 	| Date of counts 	| Date & Time 	|
| state 	| Jurisdiction 	| Plain Text 	|
| tot_cases 	| Total number of cases 	| Number 	|
| conf_cases 	| Total confirmed cases 	| Number 	|
| prob_cases 	| Total probable cases 	| Number 	|
| new_case 	| Number of new cases 	| Number 	|
| pnew_case 	| Number of new probable cases 	| Number 	|
| tot_death 	| Total number of deaths 	| Number 	|
| conf_death 	| Total number of confirmed deaths 	| Number 	|
| prob_death 	| Total number of probable deaths 	| Number 	|
| new_death 	| Number of new deaths 	| Number 	|
| pnew_death 	| Number of new probable deaths 	| Number 	|
| created_at 	| Date and time record was created 	| Date & Time 	|
| consent_cases 	| If Agree, then confirmed and probable cases are included. <br> Not Agree, then only total cases are included. 	| Plain Text 	|
| consent_deaths 	| If Agree, then confirmed and probable deaths are included. <br>If Not Agree, then only total deaths are included. 	| Plain Text 	|

## Data Questions

Questions of interest are:

**1. Find the state with the highest and lowest deaths**

**2. Compare the death rate both before and after the vaccine was released**

## Question 1
### State with highest and lowest deaths

Let's first look at this from a pure count perspective

```{r get-death-count-state}
df_tot_deaths <- df %>% 
  select(submission_date, state, tot_cases, tot_death) %>% 
  group_by(state) %>% 
  filter(submission_date == max(submission_date)) %>% 
  ungroup()

# preview
head(df_tot_deaths)
```

Since `NYC` and `NYS` are reported separately, let's isolate these values.
May need to add them together later to get a total and create a new row. 

```{r nys-andnyc}
df_tot_deaths[df_tot_deaths$state == 'NY' | df_tot_deaths$state == 'NYC',]
```

#### All States

```{r graph-states with most deaths}
df_tot_deaths %>%
  arrange(desc(tot_death)) %>% 
  ggplot(aes(x = reorder(state, tot_death), y = tot_death)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "States", y = "Total Deaths", subtitle = paste("COVID-19 Total Death Count by state as of ", Sys.Date(), sep = ""))
```

`California` has the highest count of COVID Cases.
Let's zoom in and look at the top 10.

#### Top 10

```{r top-10}
df_tot_deaths %>% 
  slice_max(tot_death, n = 10) %>% 
  ggplot(aes(reorder(state, tot_death), tot_death, fill = tot_death)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue")) +
    coord_flip() +
    theme(legend.position = 'none') +
    labs(x = "States", y = "Total Deaths", subtitle = paste("Top 10 COVID-19 Total Death Count states as of ", Sys.Date(), sep = ""))
```

Here we can see the top 10:

- California
- Texas
- Florida
- Pennsylvania
- NYC
- Ohio
- Illinois
- Georgia
- Michigan
- New Jersey

Now let's look at the bottom 10. 

#### Bottom 10

```{r bot-10}
df_tot_deaths %>% 
  slice_min(tot_death, n = 10) %>% 
    ggplot(aes(reorder(state, -tot_death), tot_death, fill = tot_death)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue")) +
    coord_flip() +
    labs(x = "States", y = "Total Deaths", subtitle = paste("Bottom 10 COVID-19 Total Death Count states as of ", Sys.Date(), sep = "")) +
    guides(fill = guide_legend(title = "Total Deaths"))
```

### Death rate by state

Now lets create a `death_rate` column and look at top_n that way.

```{r death-rate-by-state}
df_tot_deaths <- df_tot_deaths %>% 
  mutate(death_rate = tot_death / tot_cases)
```

Let's view top_n by rate.

#### Top 10 death rate

```{r top-10-rate}
df_tot_deaths %>% 
  slice_max(death_rate, n = 10) %>% 
  ggplot(aes(reorder(state, death_rate), death_rate, fill = death_rate)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue"), mid = "red") +
    coord_flip() +
    theme(legend.position = 'none') +
    labs(x = "States", y = "Death Rate", subtitle = paste("Top 10 COVID-19 Total Death Rate states as of ", Sys.Date(), sep = ""))
```

When we compare via rate, we can see California isn't even in the top 10. NYC takes the cake by almost a full 2%.

#### Bottom 10 death rate

```{r bot-10-rate}
df_tot_deaths %>% 
  slice_min(death_rate, n = 10) %>% 
    ggplot(aes(reorder(state, -death_rate), death_rate, fill = death_rate)) +
    geom_bar(stat = 'identity') +
    scale_fill_gradient2(high = muted("blue"), low = "white") +
    coord_flip() +
    labs(x = "States", y = "Death Rate", subtitle = paste("Bottom 10 COVID-19 Total Death Rate states as of ", Sys.Date(), sep = "")) +
    guides(fill = guide_legend(title = "Death Rate"))
```

### Simple US map

Now let's create a simple US map to visualize count and rate by state.
Built in package `usmaps` will only have the 52 states but is fine for our purposes.

Have to sum `NY` and `NYC` and create new row.

```{r sum-ny-nyc, warning=FALSE}
# sum total cases
NYS_tot_cases <- sum(df_tot_deaths$tot_cases[df_tot_deaths$state == "NY" | df_tot_deaths$state == "NYC"])

# sum total deaths
NYS_tot_deaths <- sum(df_tot_deaths$tot_death[df_tot_deaths$state == "NY" | df_tot_deaths$state == "NYC"])

# sum death rate
NYS_death_rate <- sum(df_tot_deaths$death_rate[df_tot_deaths$state == "NY" | df_tot_deaths$state == "NYC"])/2

#Latest date
dateLatest = as.POSIXct(max(df_tot_deaths$submission_date), origin = "1970-01-01", tz = "UTC")

# add new york summed row
df_tot_deaths[nrow(df_tot_deaths) + 1,] <- list(dateLatest, "NYS", NYS_tot_cases, NYS_tot_deaths, NYS_death_rate)

# remove separate ny and nyc
df_tot_deaths_ny <- subset(df_tot_deaths, state != "NYC" & state !="NY")

# rename NYS to NY
df_tot_deaths_ny <-  within(df_tot_deaths_ny, state[state == "NYS"] <- "NY")
```

Join `df_tot_deaths_ny` with `statepop`

```{r join-with-statepop}
df_tot_deaths_ny <-  left_join(statepop, df_tot_deaths_ny, by = c("abbr" = "state"))
```

Plot simple US map.

```{r usmap-count}
plot_usmap(data = df_tot_deaths_ny, values = "tot_death", color = "blue") +
  scale_fill_continuous(low="white", high="red", name = "Total Deaths",
                        label = scales::comma) +
  labs(subtitle = paste("COVID-19 Total Death Count by state as of ", Sys.Date(), sep = "")) +
  theme(legend.position = c(0.9,0.1))
```

Clearly California, Texas, Florida and New York have the most cases.

Let's look at this simple map via rate.

```{r usmap-rate}
plot_usmap(data = df_tot_deaths_ny, values = "death_rate", color = "blue") +
  scale_fill_continuous(low="white", high="red", name = "Total Death Rate",
                        label = scales::comma) +
  labs(subtitle = paste("COVID-19 Total Death Rate by state as of ", Sys.Date(), sep = "")) +
  theme(legend.position = c(0.9,0.1))
```

When we combine `NY` with `NYC`, we get a 0.013 death rate, therefore it does not have the highest rate. Here we can see Pennsylvania, Georga and Michigan have the darker red color and therefore higher rate. However you can visually see the difference between the count and rate US maps. Clearly they paint a different picture.

## Question 2
### Compare death rate before and after vaccine being released

According to <a href="https://www.fda.gov/news-events/press-announcements/fda-takes-key-action-fight-against-covid-19-issuing-emergency-use-authorization-first-covid-19">FDA.gov</a>, the First COVID-19 Vaccine was authorized on **Dec 11th, 2020**. This was for emergency use only, however for simplicity sake we will use this as a place marker for this analysis. 

Additionally, we will use what the World Health Organization defines as <a href="https://www.who.int/news-room/commentaries/detail/estimating-mortality-from-covid-19">Infection fatality ratio (IFR)</a> as our death rate.

This is calculated by:

$$\text{Infection fatality ratio (IFR, in %) = }\frac{\text {Number of deaths from  disease}}{\text{Number of infected  individuals}} \text{x 100}$$

Let's create two data frames, before and after the vaccination was released.

#### Before Vax data frame

```{r before-vax}
df_before_vax <- df %>% 
  select(submission_date, state, tot_death, tot_cases) %>% 
  filter(submission_date <= '2020-12-11') %>%
  group_by(state) %>% 
  filter(submission_date == max(submission_date)) %>% 
  mutate(death_rate_before = tot_death / tot_cases) %>% 
  arrange(desc(death_rate_before))

# check for NaN
df_before_vax[is.na(df_before_vax$death_rate_before),]

# preview df
head(df_before_vax)
```

Missing values are simply divide by zero error, no need to account for them using na.rm = T will work.

#### After Vax data frame

```{r after-vax}
df_after_vax <- df %>% 
  select(submission_date, state, tot_death, tot_cases) %>% 
  filter(submission_date > '2020-12-11') %>%
  group_by(state) %>% 
  filter(submission_date == max(submission_date)) %>% 
  mutate(death_rate_after = tot_death / tot_cases) %>% 
  arrange(desc(death_rate_after))

# check for NaN
df_after_vax[is.na(df_after_vax$death_rate_after),]

# preview df
head(df_after_vax)
```

No missing values in `df_after_vax`


Let's create an overall data frame to get a complete picutre of how the release of the vaccine effected death rates.

#### Overall data frame

```{r join-on-state}
df_death_rate <- df_before_vax %>% 
  left_join(df_after_vax, by = "state") %>% 
  select(state, death_rate_before, death_rate_after)

head(df_death_rate)
```

What is the total death rate before and after vaccine?

```{r total-death-rate-before-after}
# death rate before
before_vax <- sum(df_death_rate$death_rate_before, na.rm = T)/nrow(df_death_rate)

# death rate after
after_vax <- sum(df_death_rate$death_rate_after, na.rm = T)/nrow(df_death_rate)

# create new df
df_death_rate_sum <- data.frame(vax_status = c("Before Vax", "After Vax"), 
                                death_rate = c(before_vax, after_vax))

df_death_rate_sum
```

Let's visualize this table.

```{r simple-death-rate-sum}
ggplot(df_death_rate_sum, aes(x=death_rate, y= vax_status, fill = vax_status)) +
  geom_bar(stat = 'identity') +
  labs(x = "Death Rate", y = "", subtitle = paste("Overall COVID-19 US Death Rate as of ", Sys.Date(),sep = "")) +
  theme(legend.position = "none")
```

## Conclusion

### Question 1

**Which states had the most and least deaths?**

- While California had the most deaths, the death rate (total cases / total deaths) was highest in NYC
- Least deaths in the continental United States was Vermont with ~600, they also were in the *bottom 10* for lowest death rate. Utah was first in lowest death rate with ~0.0045.

### Question 2

**What was the death rate before and after the vaccine was released?**

- If we use December 11th 2020 as the after vaccination date, the aggregate death rate for the entire data set is 0.016 before the vaccine was released, and 0.010 after. So we see a pretty significant decrease in total death rate after the vaccine was released. 







