---
title: "Sentiment Analysis"
author: "Justin Williams"
date: "4/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyselect)
library(tidyverse)
library(pdftools)
library(stringr)
```

## Overview

This is a demostration of using sentiment analysis to analyze a corpus.

### Load Corpus

I found a copy of my favorite book in `pdf` online, so I will download it and convert it to csv for sentiment analysis.

```{r convert-pdf-csv}
# pdf url
url <- "https://ia802800.us.archive.org/4/items/JohnRobbinsDietForANewAmerica/John_Robbins_Diet_for_a_New_America.pdf"

# download file
download.file(url, destfile = basename(url), method="curl", extra="-k")

#load pdf
diet_new_am <- 
  pdf_text("John_Robbins_Diet_for_a_New_America.pdf") %>% 
  str_split("\n")

# check type
class(diet_new_am)

# how many pages
length(diet_new_am)
```

Now it's a list with 463 nested lists. Let's separate each lines characters.

```{r split-lines}
# squish then split
diet_new_am <- diet_new_am %>% 
  str_squish() %>% # squish lines
  strsplit(split= "\\,\\s\\\"") #split on spaces


# get rid of "c("
for(i in 1:length(diet_new_am)) {
   diet_new_am[[i]][1] <- diet_new_am[[i]][1] %>%
    str_extract("(?<=c[:punct:]\\\").*")
}

# get rid of \
for(i in 1:length(diet_new_am)) {
  for(j in 1:length(diet_new_am[[i]])) {
    diet_new_am[[i]][j] <- diet_new_am[[i]][j] %>%
      str_extract(".*(?=\")")
    }
}

head(diet_new_am)
diet_new_am[[5]][1]
```


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
