---
title: "Assignment 2 - R and SQL"
author: "Justin Williams"
date: "2/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE}
library(RPostgres)
library(DBI)
```

## Connect to PostgreSQL Database

Database username and password will be prompted by `askForPassword` function from rstudioapi.

```{r connect-database}
conn <- dbConnect(
  RPostgres::Postgres(),
  dbname = "movies",
  host = "localhost",
  port = "5434",
  user = rstudioapi::askForPassword("Database user"),
  password = rstudioapi::askForPassword("Database password")
)
```


Let's verify that the three tables exist in the database

```{r verify-tables}
dbExistsTable(conn,"people")
dbExistsTable(conn,"movies")
dbExistsTable(conn,"ratings")
```

## Read tables in R
Once verified, we can read our tables into R using dbReadTable:

```{r read-tables}
people <- dbReadTable(conn,c("people"))
movies <- dbReadTable(conn, c("movies"))
ratings <- dbReadTable(conn,c("ratings"))
```

Take a look at the tables:
```{r}
people
movies
ratings
```
## Editing database directly from R

I surveyed my friends to see what their ratings were on the aforementioned movies, so now I need to update the `ratings` table with the results.

```{r edit-database}
res <- dbSendQuery(conn, 
                   statement = 
                     paste("INSERT INTO ratings (person_id, movie_id, rating) VALUES   ('1','1','4'),('1','2',NULL),('1','3','2'),('1','4','3'),('1','5','4'),('1','6','1'),
('2','1','4'),('2','2',NULL),('2','3','3'),('2','4','3'),('2','5',NULL),('2','6',NULL),
('3','1','3'),('3','2','3'),('3','3',NULL),('3','4',NULL),('3','5','3'),('3','6',NULL),
('4','1','4'),('4','2','3'),('4','3',NULL),('4','4','5'),('4','5','4'),('4','6',NULL),
('5','1','5'),('5','2',NULL),('5','3','3'),('5','4',NULL),('5','5',NULL),('5','6','4')"));
```

Look at updated ratings table:
```{r updated-ratings}
ratings <- dbReadTable(conn,c("ratings")) # read updated table in from db
ratings
```

