---
title: "Assignment 1 - Loading Data into a Data Frame"
author: "Justin Williams"
date: "1/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

## What Do Men Think It Means To Be A Man?

### Overview

The article was essentially about men's idea about masculinity, and its subsequent role in society. A survey was given in light of the `#MeToo` movement, which is/was a movement where folks came out about their abuses primarily from men. The goal of this survey was to attempt to quantify if the `#MeToo` movement had any effect on how men feel. Did it change their ideas of what makes a man a man, or has it fallen on deaf ears.
<br>
<br>
The article is linked here:  
<a href="https://fivethirtyeight.com/features/what-do-men-think-it-means-to-be-a-man">What Do Men Think It Means To Be A Man?</a>.

```{r}
library(readr)
library(dplyr)
menSurvey <- read_csv("/Users/justinwilliams/Documents/CUNY SPS/607/Assignments/1/masculinity-survey/raw-responses.csv", show_col_types = F,col_select = 2:98)
head(menSurvey)
```

### Data Transformations
- Columns for `StartDate` and `EndDate` were kept as record indicators
- Question 5:  
  *Do you think that society puts pressure on men in a way that is unhealthy or bad for them?* <br>Yes/No answers, were changed to 1, 0 or NA and datatype was altered to numeric to enable summing.
- Columns with *Not Selected* or *No answer* where changed to NA.
- Most demographic columns were kept and given more descriptive names.
- A subset of what I felt the most interesting questions where kept.

All of this is outlined in the tables below

#### Demographic info

| Demographic 	| Column Name 	| New Column Name 	|
|---	|---	|---	|
| Race 	| race2 	| Race 	|
| Ethnicity 	| racethn4 	| Ethnicity 	|
| Age 	| age3 	| Age 	|
| Sexual Orientation 	| orientation 	| Orientation 	|
| Education 	| educ4 	| Education Level 	|

#### Subset of questions

| Question                                                                                                                   	| Column Name 	| New Column Name 	|
|----------------------------------------------------------------------------------------------------------------------------	|-------------	|-----------------	|
| In general, how masculine or "manly" do you feel?                                                                          	| q0001       	| Question_1      	|
| How important is it to you that others see you as masculine?                                                               	| q0002       	| Question_2      	|
| Do you think that society puts pressure on men in a way that is unhealthy or bad<br>for them?                              	| q0005       	| Question_5      	|
| How often would you say you do each of the following?  Cry                                                                 	| q0007_0004  	| Question_7_4    	|
| How often would you say you do each of the following? Watch sports of any kind                                             	| q0007_0008  	| Question_7_8    	|
| How often would you say you do each of the following? See a therapist                                                      	| q0007_0010  	| Question_7_10   	|
| How often would you say you do each of the following? Feel lonely or isolated                                              	| q0007_0011  	| Question_7_11   	|
| Which of the following do you worry about on a daily or near daily basis? (Select<br>all that apply.) Your mental health   	| q0008_0008  	| Question_8_8    	|
| Which of the following do you worry about on a daily or near daily basis? (Select<br>all that apply.) Your physical health 	| q0008_0009  	| Question_8_9    	|

```{r}
# subset and rename columns
menSurveySubset <- menSurvey %>% select(StartDate, EndDate, Age=age3, Race=race2, Ethnicity=racethn4, Education=educ4, Sexual_Orientation=orientation, Question_1=q0001, Question_2=q0002, Question_5=q0005, Question_7_4=q0007_0004, Question_7_8=q0007_0008,  Question_7_10=q0007_0010, Question_7_11=q0007_0011, Question_8_8=q0008_0008, Question_8_9=q0008_0009)

head(menSurveySubset)
```
```{r}
# change Yes/No/No Answer to 1/0 or NA
menSurveySubset$Question_5[menSurveySubset$Question_5=="Yes"] <- 1
menSurveySubset$Question_5[menSurveySubset$Question_5=="No"] <- 0
menSurveySubset$Question_5[menSurveySubset$Question_5=="No answer"] <- NA
```

```{r}
# Convert Question 5 to numeric
menSurveySubset$Question_5 <- as.numeric(menSurveySubset$Question_5)
```

```{r}
# change any 'Not Selected' values within df to NA
menSurveySubset[(menSurveySubset == "Not selected") | (menSurveySubset == "No answer")] <- NA
```

```{r}
# view results of transformations
head(menSurveySubset[,c("Sexual_Orientation", "Question_5", "Question_8_8")])
```
### Conclusion
The article concludes suggesting that American men are still fractured about their role in society, and that despite a years worth of the `#MeToo` headline, it does not mean the majority of men have altered their views on masculinity. For next steps, I think an update to this survey (its from 2018) would be appropriate. The `#MeToo` movement has continued to remain in the limelight, and if administered again, we would perhaps see significantly different survey results.





