---
title: "Assignment 5 - Tidying and Transforming Vaccination Data"
author: "Justin Williams"
date: "3/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results=FALSE}
library(tidyverse)
library(readxl)
```

## Overview

We were asked to analyze data from August 2021 Israeli hospitalization ("Severe Cases") rates for pople under 50(assume "50 and under") and over 50, for both un-vaccinated and fully faccinated populations. Specific questions are as follows:
*1) Do you have enough information to calculate the total population? What does this total population represent?*
*2) Calculate the Efficacy vs. Disease; Explain your results.*
*3) From your calculation of efficacy vs. disease, are you able to compare the rate of severe cases in unvaccinated individuals to that in vaccinated individuals?*

<u>**Definitions**</u>
Severe Cases = hospitatlized
Efficacy vs. sever disease = 1 - (% fully vaxed sever cases per 100k / % not vaxed sever cases per 100k)

### Additional domain knowledge needed
1) Israel's total population
According to <a href="https://en.wikipedia.org/wiki/Israel">Wikipedia</a> the 2022 population estimate is 9,481,820

2) Who is eligible to receive vaccinations

According to an article in <a href="https://www.usnews.com/news/health-news/articles/2021-02-04/all-israelis-over-16-are-eligible-for-coronavirus-vaccine">US News</a> as of Feb 4th 2021 everyone over the age of 16 is eligivle to receive a vaccine in Israel.

3) What does it mean to be fully vaccinated

## Read in data

```{r load-data}
read_excel("./data/israeli_vaccination_data_analysis_start.xlsx",
           range = "A1:F6")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
